---
title: "ADDS_project"
output: html_document
date: "2023-04-10"
---


##COADDS22.1F--017

#Reading the data set
```{r}
library(dplyr)
library(ggplot2)
employee_raw_data = read.csv("HR_Analytics.csv")
View(employee_raw_data )
```

#Extracting the using variables
```{r}
employee_new_data <- data.frame(employee_raw_data$JobSatisfaction,employee_raw_data$Attrition,employee_raw_data$OverTime,employee_raw_data$EnvironmentSatisfaction,employee_raw_data$HourlyRate,employee_raw_data$Education,employee_raw_data$WorkLifeBalance)

View(employee_new_data)

```

#Visualization the variables seperately
```{r}
#Contingency table for attrition
table1 <- table(employee_new_data$employee_raw_data.Attrition)
table1
```
```{r}
# Plot the bar chart
barplot(table1,beside = F,
        col=rainbow(2),
        legend=rownames(table1),
        main="Barplot of Attrition Variable",
        xlab="Attrition",
        ylab="frequency")
abline(h=0)

```

## Filter the dataframe to include only employees who have left the organization

```{r}
left_df <- subset(employee_new_data,employee_new_data$employee_raw_data.Attrition == "Yes")
```

## Counting the number of occurrences of each education level of the employees who left the organization.
```{r}
edu_counts = table(left_df$employee_raw_data.Education)
edu_counts
```
#Plotting pie chart to identify the level of education of the employees who left the organization
```{r}
catego <- c("Below college", "College", "Bachelor", "Master","Doctor")
val <- c(31,44,99,58,5)
pie(val, labels = catego, main = "Pie Chart for educational level")


```


```{r}
#Contingency table for Overtime
table2 <- table(employee_new_data$employee_raw_data.OverTime)
table2
```
#Creating Bar plot for Over time variable
```{r}
# Plot the bar chart
barplot(table2,beside = F,
        col=rainbow(2),
        legend=rownames(table2),
        main="Barplot of Overtime Variable",
        xlab="Overtime",
        ylab="frequency")
abline(h=0)
```

```{r}
#Create histogram for HourlyRate
hist(employee_new_data$employee_raw_data.HourlyRate,col="lightblue", 
     main = "Histogram of HourlyRate")
abline(h=0)
```

```{r}
#Contingency table for job satisfaction
table3 <- table(employee_new_data$employee_raw_data.JobSatisfaction)
table3
```
```{r}
# Plotting the bar chart for Job satisfaction variable
barplot_data2 <- barplot(table3, beside = T,
                         col = rainbow(4),
                         legend = rownames(table3),
                         main = "Barplot of Job satisfaction Variable",
                         xlab = "Level of Job Satisfaction",
                         ylab = "frequency",
                         ylim = c(0, max(table3) + 20)) # Extend the y-axis range
abline(h = 0)


text(x = barplot_data2, y = table3 + 2, label = table3, cex = 1, col = "black")

```
```{r}
#Contingency table for environment satisfaction
table4 <- table(employee_new_data$employee_raw_data.EnvironmentSatisfaction)
table4
```
```{r}
# Plotting the bar chart for environment satisfaction variable
barplot_data3 <- barplot(table4, beside = F,
                         col = c("#FF6666", "#FFCC66", "#66FF66", "#66CCFF"),
                         legend = rownames(table4),
                         main = "Barplot of environment satisfaction Variable",
                         xlab = "Level of environment satisfaction",
                         ylab = "frequency",
                         ylim = c(0, max(table4) + 20)) # Extend the y-axis range
abline(h = 0)


text(x = barplot_data3, y = table4 + 2, label = table4, cex = 1, col = "black")

```
```{r}
#Contingency table for work-life balance
table5 <- table(employee_new_data$employee_raw_data.WorkLifeBalance)
table5
```
```{r}
# Plotting the bar chart for work-life balance
barplot_data <- barplot(table5, beside = F,
                        col = rainbow(4),
                        legend = names(table5),
                        main = "Barplot of work-life balance",
                        xlab = "work-life balance",
                        ylab = "frequency",
                        ylim = c(0, max(table5) + 20)) # Extend the y-axis range
abline(h = 0)

# Adding data labels to the bar plot
text(x = barplot_data, y = table5 + 2, label = table5, cex = 1, col = "black")




```

#Main Objectives

#Main Objective1 - Identifying is there an impact on employee attrition from the selected job-related factors.

#Sub objective 1 - Identifying is there a relationship between job satisfaction and attrition.

```{r}
#Here dependent variable is attrition and the independent variable is job satisfaction

#Null hypothesis (H0) -There are no relationship between job satisfaction and attrition. 
#Alternative hypothesis (H1) - -	There is a relationship between job satisfaction and attrition. 

table <- table(employee_new_data$employee_raw_data.JobSatisfaction,employee_new_data$employee_raw_data.Attrition)
table

test <- chisq.test(table)

test

```
```{r}
#For this, p-value is less than 0.05. So we can reject null hypothesis. Which means,There is a relationship between job satisfaction and attrition variables.
```
#Visualizing the output using a heatmap.

```{r}
ggplot(as.data.frame(table), aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Job Satisfaction", y = "Attrition", fill = "Frequency") +
  ggtitle("Heatmap of Job Satisfaction vs Attrition")
```
```{r}
#In here, Cells with higher frequencies will be represented by darker colors on the heatmap, while cells with lower frequencies will be lighter.
#Here the cell corresponding to high Job Satisfaction and Attrition = No has a high frequency, it means that many employees with high job satisfaction did not leave the company.The cell corresponding to low Job Satisfaction and Attrition = no has a low frequency, it means that only fewer employees with low job satisfaction did not leave the company.

```


#Sub objective 2 - Identifying is there a relationship between working overtime and attrition

```{r}
#Here dependent variable is attrition and the independent variable is working overtime

#Null hypothesis (H0) -There are no relationship between overtime and attrition. 
#Alternative hypothesis (H1) - -	There is a relationship between overtime and attrition. 

table1 <- table(employee_new_data$employee_raw_data.OverTime,employee_new_data$employee_raw_data.Attrition)
table1

test1 <- chisq.test(table1)

test1
```

```{r}
#For this, p-value is less than 0.05. So we can reject null hypothesis. Which means,There is a relationship between working overtime and attrition variables.
```
#Visualizing the output using a heatmap.

```{r}
ggplot(as.data.frame(table1), aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Working Overtime", y = "Attrition", fill = "Frequency") +
  ggtitle("Heatmap of Working Overtime vs Attrition")

```
```{r}
#A high frequency in a cell indicates that there are many occurrences of that specific combination of Working Overtime and Attrition levels. The cell corresponding to "Working Overtime = No" and "Attrition = No" has a high frequency, it means that many employees who did not work overtime don't have attrition which means they did not leave the company. In other word we can say that,

#A low frequency in a cell indicates that there are fewer occurrences of that specific combination of Working Overtime and Attrition levels. The cell corresponding to "Working Overtime = Yes" and "Attrition = No" has a low frequency, it means that there are only fewer employees who worked overtime did not leave the company
```

#Sub objective 3 - Identifying is there a relationship between Environment satisfaction and attrition
```{r}
#Here dependent variable is attrition and the independent variable is environment satisfaction variable

#Null hypothesis (H0) -There are no relationship between environment satisfaction and attrition. 
#Alternative hypothesis (H1) - -	There is a relationship between environment satisfaction and attrition.

table2 <- table(employee_new_data$employee_raw_data.EnvironmentSatisfaction,employee_new_data$employee_raw_data.Attrition)
table2

test2 <- chisq.test(table2)

test2
```

```{r}
#For this, p-value is less than 0.05. So we can reject null hypothesis. Which means,There is a relationship between envirionment satisfaction and attrition variables.
```
#Visualizing the output using heatmap

```{r}
ggplot(as.data.frame(table2), aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Environment Satisfaction", y = "Attrition", fill = "Frequency") +
  ggtitle("Heatmap of Environment Satisfaction vs Attrition")

```
```{r}
#A high frequency in a cell indicates that there are many occurrences of that specific combination of Environment Satisfaction level and Attrition status. The cell corresponding to "Environment Satisfaction = 3 and 4" and "Attrition = No" has a high frequency, it means that many employees with a level 3 and 4 environment satisfaction did not leave the company(They do not have attrition)
#Inother word, A low frequency in a cell indicates that there are fewer occurrences of that specific combination of Environment Satisfaction level and Attrition status. The cell corresponding to "Environment Satisfaction = 2 and 1" and "Attrition = NO" has a low frequency, it means that fewer employees with a level 2 and 1 environment satisfaction are staying in the company which means so many left the company.
```


#Sub objective 4 - Identifying is there a relationship between Hourly rate and attrition
```{r}
#Here dependent variable is attrition and the independent variable is Hourly rate of pay variable

#Null hypothesis (H0) -There are no relationship between Hourly rate and attrition. 
#Alternative hypothesis (H1) - -	There is a relationship between Hourly rate of and attrition.
# View unique values in the dependent variable
unique(employee_new_data$employee_raw_data.Attrition)

# Convert "Yes" to 1 and "No" to 0
employee_new_data$employee_raw_data.Attrition <- as.numeric(employee_new_data$employee_raw_data.Attrition == "Yes")

# Check the updated unique values
unique(employee_new_data$employee_raw_data.Attrition)




```
```{r}
# Fit logistic regression model
model <- glm(employee_new_data$employee_raw_data.Attrition ~ employee_new_data$employee_raw_data.HourlyRate, data = employee_raw_data, family = "binomial")

# Summarize the model
summary(model)

```

```{r}
#Based on the output of the logistic regression model, we have a non-significant p-value (0.793), which suggests that there is a relationship between Hourly Rate and Attrition.
```
#Visualizing the output using boxplot
```{r}
ggplot(employee_new_data1, aes(x = employee_raw_data.Attrition, y = employee_raw_data.HourlyRate)) +
  geom_boxplot() +
  labs(x = "Attrition", y = "Hourly Rate") +
  ggtitle("Boxplot of Hourly Rate vs Attrition")

```

#Since the Logistic regression test suggests no significant relationship,  we do not have substantial differences in the boxplot. 


#Main objective 3 - Evaluate is there a relationship between an employee's perception of their work-life balance and their attrition.
```{r}
#Here dependent variable is attrition and the independent variable is work-life balance variable

#Null hypothesis (H0) -There are no relationship between work-life balance variable and attrition. 
#Alternative hypothesis (H1) - -	There is a relationship between work-life balance variable and attrition.

table6 <- table(employee_new_data$employee_raw_data.WorkLifeBalance,employee_new_data$employee_raw_data.Attrition)
table6

test6 <- chisq.test(table6)

test6
```
```{r}
#For this, p-value is less than 0.05. So we can reject null hypothesis. Which means,There is a relationship between employees perception about their work-life balance and attrition variables.
```
#Visualizing the output using heatmap

```{r}
ggplot(as.data.frame(table6), aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Work-Life Balance", y = "Attrition", fill = "Frequency") +
  ggtitle("Heatmap of Work-Life Balance vs Attrition")

```
```{r}
#A high frequency in a cell indicates that there are many occurrences of that specific combination of Work-Life Balance level and Attrition status. The cell corresponding to "Work-Life Balance = 3" and "Attrition = No" has a high frequency, it means that many employees with a level 3 work-life balance did not leave the company (they did not have attrition).

#A low frequency in a cell indicates that there are fewer occurrences of that specific combination of Work-Life Balance level and Attrition status. For example, if the cell corresponding to "Work-Life Balance = 1" and "Attrition = Yes" has a low frequency, it means that fewer employees with a level 1 work-life balance left the company (experienced attrition).
```
```{r}
R.version

```

